---
title: "Lesson 2: Math Operators"
---

### Use R as a calculator:

Addition, substraction, multiplication, division
```{r}
1 + 3
10 - 15
2*8
60/12
```

exponentiate
```{r}
3^2
2^4
9^0
2^-2
```

square root
```{r}
sqrt(4)
```

logs
```{r}
log(10)
```
(base e)

log with base 10

```{r}
log10
```

### Order of operations.

What results from this?
```{r}
2*6 - 6*9 + 3*30 /7
```

The rules:  

1. operations go left to right
2. multiplication and division before substraction and/or Addition
....

There's more: 

integer division (the remainder is discarded)
```{r}
5 %%/ 3
```

modulus operator (return the remainder after division)
```{r}
5/3
```
....and so much more 

### Some notes on R syntax

- most often, the amount of white space does not matter. 
- These are the same:

```{r}
4/3
4/    3
4    /  3
```

- these are also the same
```{r}
log(10)
log( 10 )
log ( 10)

- a hard return between lines of code is enough to separate the commands
- exception: unclosed parentheses, brackets, or quotes. R will wait for these to be completed. A single quote must always be complemented by a second single quote, and a double quote likewise must always have a second quote to complete it. Left parentheses, curly braces, or brackets much also be accompanying by their right-sided complement. 
- good examples
```
{ }
( )
[ ]
"  "
' '
` `
```

- bad examples
```
(  } ] ' "
```

- there is no difference between double and single quotes on a practical level, but R will interpret them as different commands (so a single quote cannot close a double quote). This is useful when there is nested levels of quoting (rare), and yet, it happens now and then. 

