---
title: "Getting to Know Your Data in R"
editor: 
  markdown: 
    wrap: 72
---


::: {.callout-caution collapse="false"}
## Learning Goals

At the end of this lesson, you should:

-   How to import data in R
-   know several methods of data exploration in R
-   understand how to make cross tabulations of data
-   know how to make histograms and pairwise plots in R
:::

### Import a csv data set

This code ("read_csv") reads a CSV file named "trial_data.csv" and assigns it to a variable named "data1".
The second line of code uses the "head" function to display the first 5 rows of the "data1" data frame.
```{r, eval=FALSE}
data1 <- read.csv("trial_data.csv")

head(data1, 5)
view(data1)
```

```{r, echo=FALSE}
data1 <- read.csv(here::here("data/trial_data.csv"))
head(data1, 5)
#View(data1)
```
### Explore data

Base R also has some useful functions for quickly exploring dataframes:

-   str: Show the structure of an R object, including a dataframe
-   summary: Give summaries of each column of a dataframe

```{r}
str(data1)

summary(data1)
```


### Make Some Simple Plots

### Calculate Some Values

### Access Data

::: callout-tip
## Putting it all together

?????
:::
