---
title: "Getting to Know Your Data in R"
editor: 
  markdown: 
    wrap: 72
---


::: {.callout-caution collapse="false"}
## Learning Goals

At the end of this lesson, you should:

-   How to import data in R
-   know several methods of data exploration in R
-   understand how to make cross tabulations of data
-   know how to make histograms and pairwise plots in R
:::

### Import a csv data set

This code ("read_csv") reads a CSV file named "trial_data.csv" and assigns it to a variable named "data1".
The second line of code uses the "head" function to display the first 5 rows of the "data1" data frame. 
The view() will open the data set in new RStuido video where you can look up at the rows and columns. 
```{r, eval=FALSE}
data1 <- read.csv("trial_data.csv")

head(data1, 5)
view(data1)
```

```{r, echo=FALSE}
data1 <- read.csv(here::here("data/trial_data.csv"))
head(data1, 5)
#View(data1)
```
### Explore data

Base R also has some useful functions for quickly exploring dataframes:

-   str: Show the structure of an R object, including a dataframe
-   summary: Give summaries of each column of a dataframe

```{r}
str(data1)

summary(data1)
```


### Exploring Data in R 

To access the data in any of the variables (columns) in our data frame we can use a $ notation. For example, to access the 'variety' variable in our data1 data frame we can use the code below. This tells R that the variety variable is contained within the data frame data1.

```{r eval=FALSE}
data1$variety

```


### Make Some Simple Plots

Here we are creating a histogram to look at data distribution of the 'yield' variable from data1 data frame using a hist() function.

A boxplot() function in R is used to create a boxplot for the selected variables. In the code chunk below, a boxplot of yield for each replication is createdfrom data1, the xlab and ylab shows the title of x-axis and y-axis, respectively. The main=, gives the title to the graph.

```{r}
hist(data1$yield)


# Create the box plot
boxplot(yield ~ rep, data = data1,
        main = "Yield Graph",
        xlab = "Rep",
        ylab = "Yield")


```


### Calculate Some Values
```{r}


```


### Access Data

::: callout-tip
## Putting it all together

?????
:::
