---
title: "Math Operators"
---

::: {.callout-caution collapse="false"}
## Learning Goals

At the end of this lesson, you should:

-   Understand sorts of math operators available in R and how to use them
-   Understand logical operators in R
-   Be aware of the order of operations
-   Be aware of how whitespace is interpreted in R
-   Understand how to use parentheses, brackets, braces and quotations in R
:::

### R as a calculator

#### Addition, subtraction, multiplication, division

```{r}
1 + 3
10 - 15
2*8
60/12
```

A hard return between lines of code is sufficient to separate the commands.

#### Exponentiation

```{r}
3^2
2^4
9^0
2^-2
```

R can also handle scientific notation. This number, `3e2` is equivalent to $3 * 10^2$, or $3000$.

#### Roots (square, cube, ....)

```{r}
4^(1/2)
8^(1/3)
```

#### Logs

```{r}
log(10)
```

(base e)

log with base 10

```{r}
log10(10)
```

log with base 2

```{r}
log2(4)
```

If you have other bases:

```{r}
log(10, base = 4)
```

#### Operations with sign

*(positive and negative signs are called "unary operators")*

```{r}
3*-4
```

Like in standard math, only negatively signed numbers need to be specified.

#### Integer division (the remainder is discarded)

```{r}
5 %/% 3
```

#### Modulus operator (return the remainder after division)

```{r}
5 %% 3
```

....and so much [more](https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html)

### Logical Operators

These test for conditions ("is this true?") and return either a `TRUE` or `FALSE`

| syntax   | Function                                            |
|----------|-----------------------------------------------------|
| ==       | equal                                               |
| !=       | does not equal                                      |
| \<       | less than                                           |
| \>       | greater than                                        |
| \<=, \>= | less than and equal to, and greater than equivalent |

**Examples**

```{r}
1 == 1
1 == 2
1 != 2
1 < 1
1 > 1
1 <= 1
1 >= 1
```

When testing multiple conditions: use `&` ('and') if two things must be true and `|` ('or') if one of two things must be true:

```{r}
1 < 2 & 1 != 1 
1 < 2 | 1 != 1
```

### Order of operations.

The rules:

1.  operations go left to right
2.  exponents are first, followed by 'unary operators' (+/- signs)
3.  multiplication and division before subtraction and/or addition
4.  logical operators come after all mathematical transformations
5.  Parentheses overall all other rules!

What results from this?

```{r, eval=FALSE}
2^3+4+12*7/2 <= -6*9
```

**When in doubt about the order of operations use parentheses!**

[Here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Syntax.html) is the official R guide to order of operations (warning: this is complicated and refers to functions beyond mathematical operators).

### Some notes on R syntax

-   most often, the amount of white space does not matter.

These are the same:

```{r}
4/3
4/    3
4    /  3
```

These are also the same:

```{r}
log(10)
log( 10 )
log ( 10)
```

-   R expects certain things to be paired or completed before it will send it to the interpreter

-   As mentioned, earlier a hard return is sufficient to send a command to the R interpreter.

-   Exceptions: binary operators (= those expecting 2 numbers): `+`, `-`, `*`, `/`, `^`, `==`, etc

-   Exceptions: unclosed parentheses `()`, brackets `[]` `{}`, or quotes `' '` `" "`. R will wait for these to be completed. A single quote must always be complemented by a second single quote, and a double quote likewise must always have a second quote to complete it. Left parentheses, curly braces, or brackets much also be accompanying by their right-sided complement.

-   good examples

<!-- -->

    1 + 2
    { }
    ( )
    [ ]
    "  "
    ' '
    ` `

-   bad examples

<!-- -->

    1 + 
    '
    (  } ] 
    "
    ' "

-   there is no difference between double and single quotes on a practical level, but R will interpret them as different commands (so a single quote cannot close a double quote). This is useful when there is nested levels of quoting (rare), and yet, it happens now and then.

::: callout-note
If you become stuck with an unfinished command, you can use the escape key, `ESC`, to get out of it.
:::

::: callout-tip
## Putting it all together

Check the "History" tab in the upper right hand pane (this should be to the right of the "Environment" tab). What is there?

If you followed along and coded the above examples, you should see the command you ran previously (including any mistakes). This is your command history. There are several icons directly above your history - explore what those do (hoover before clicking any icon to make sure you are okay with action before performing it).
:::
