---
title: "Lesson 11: Combining Data Sets"
---

::: {.callout-caution collapse="false"}
## Learning Goals

At the end of this lesson, you should: 

* understand the concept of a "key" for merging
* be able to merge two data sets together
* know the difference beween left join, right join, full join, semi-join and anti-join
:::

```{r, include=FALSE}
library(dplyr)
```


## Bind rows

Data sets are assumed to have a single observation in each row, and each column is a variable describing each observation in some way. 

If you have two data sets of different observations and the same columns, you can do a row bind to stack the data sets:

```{r, eval=FALSE}
new1 <- bind_rows(x, y)
```

The column names are match and ordered according to the first data frame listed. The default behavior is to return all unique columns from both data sets and fill in with missing data as needed. 

## Joins

Merging two data sets when it goes beyond a row bind can take an effort. 

All joins follow this syntax: 

```{r, eval=FALSE}
xxx_join(left_dataset, right_dataset)
```

"left" and "right" correspond to the left and right ....


### Left join

Return a merged data set with all the observations

### Right join

### Full join

All observations are returned, regardless if matched. 

### inner join

### 

Only observations found in both data sets are kept. 

## Final Thoughts

::: {.callout-tip}
## Putting it all together


:::

